MRD_final_microbiome

### 2019-06-25 import_popln_data.Rmd
updated

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/globalmicrobiome')
# test
#test
#test
#clash 1
#clash2maryam

```{r functions}

##### FUNCTIONS ########
##### Normalization #######

# Better rounding function than R's base round
myround <- function(x) { trunc(x + 0.5) }

# Scales reads by 
# 1) taking proportions
# 2) multiplying by a given library size of n
# 3) rounding 
# Default for n is the minimum sample size in your library
# Default for round is floor
scale_reads <- function(physeq, n = min(sample_sums(physeq)), round = "floor") {

  # transform counts to n
  physeq.scale <- transform_sample_counts(physeq, 
    function(x) {(n * x/sum(x))}
  )

  # Pick the rounding functions
  if (round == "floor"){
    otu_table(physeq.scale) <- floor(otu_table(physeq.scale))
  } else if (round == "round"){
    otu_table(physeq.scale) <- myround(otu_table(physeq.scale))
  }

  # Prune taxa and return new phyloseq object
  physeq.scale <- prune_taxa(taxa_sums(physeq.scale) > 0, physeq.scale)
  return(physeq.scale)
}

##### ADONIS ###########

# Function to run adonis test on a phyloseq object and a variable from metadata
# Make sure OTU data is standardized/normalized before 
phyloseq_to_adonis <- function(physeq, distmat = NULL, dist = "bray", formula) {

  if(!is.null(distmat)) {
    phydist <- distmat
  } else {
    phydist <- phyloseq::distance(physeq, dist)
  }

  metadata <- as(sample_data(physeq), "data.frame")

  # Adonis test
  f <- reformulate(formula, response = "phydist")
  adonis.test <- adonis(f, data = metadata)
  print(adonis.test)

  # Run homogeneity of dispersion test if there is only 1 variable
  if (grepl("\\+", formula)) {
    l <- list(
      dist = phydist, 
      formula = f, 
      adonis = adonis.test
    )
  } else {
    group <- metadata[ ,formula]
    beta <- betadisper(phydist, group)
    disper.test = permutest(beta)
    print(disper.test)

    l <- list(
      dist = phydist, 
      formula = f, 
      adonis = adonis.test, 
      disper = disper.test
    )
  }
  return (l)
}

########## Bar Plots #################

# This function takes a phyloseq object, agglomerates OTUs to the desired taxonomic rank, 
# prunes out OTUs below a certain relative proportion in a sample (ie 1% ) 
# and melts the phyloseq object into long format which is suitable for ggplot stacked barplots.
taxglom_and_melt <- function(physeq, taxrank, prune){

  # Agglomerate all otu's by given taxonomic level
  pglom <- tax_glom(physeq, taxrank = taxrank)

  # Create a new phyloseq object which removes taxa from each sample below the prune parameter
  pglom_prune <- transform_sample_counts(pglom,function(x) {x/sum(x)})
  otu_table(pglom_prune)[otu_table(pglom_prune) < prune] <- 0
  pglom_prune <- prune_taxa(taxa_sums(pglom_prune) > 0, pglom_prune)

  # Melt into long format and sort by taxonomy
  physeq_long <- psmelt(pglom_prune)
  physeq_long <- physeq_long[order(physeq_long[ ,taxrank]), ]

  # Return long data frame
  return(physeq_long)
}

###### Merge functions ############

# Merge samples by averaging OTU counts instead of summing
merge_samples_mean <- function(physeq, group, round){
  # Calculate the number of samples in each group
  group_sums <- as.matrix(table(sample_data(physeq)[ ,group]))[,1]

  # Merge samples by summing
  merged <- merge_samples(physeq, group)

  # Divide summed OTU counts by number of samples in each group to get mean
  # Calculation is done while taxa are columns
  x <- as.matrix(otu_table(merged))
  if(taxa_are_rows(merged)){ x<-t(x) }

  # Pick the rounding functions
  if (round == "floor"){
    out <- floor(t(x/group_sums))
  } else if (round == "round"){
    out <- myround(t(x/group_sums))
  }

  # Return new phyloseq object with taxa as rows
  out <- otu_table(out, taxa_are_rows = TRUE)
  otu_table(merged) <- out
  return(merged)
}

# Merge samples, just including OTUs that were present in all merged samples
# Call this function before running merge_samples()
merge_OTU_intersect <- function(physeq, group){

  # Make sure we're not starting with more taxa than we need 
  physeq <- prune_taxa(taxa_sums(physeq) > 0, physeq)

  s <- data.frame(sample_data(physeq))
  l <- levels(s[,group])
  o <- otu_table(physeq)

  # Loop through each category
  for (cat in 1:length(l)) {

    # Get the index of all samples in that category
    w <- which(s[,group]==l[cat])

    # subset to just those columns of OTU table
    cat.sub<-o[,w]
    print(dim(cat.sub))

    # Find the indices of 0's in the OTU table
    zeros <- apply(cat.sub, 1, function(r) any(r == 0))

    # If an OTU had a 0 in at least one sample, change all samples to 0
    cat.sub[zeros,] <- 0
  }

  o[,w] <- cat.sub
  otu_table(physeq) <- o

  return(physeq)

}

@@

pseq <- baseline(atlas1006) # only uses the baseline samples. This is n=1006

@@

## MND,PD,AD Variable
```{r mnd variable}

# Create mnd variable
# grabs the nationality from phyloseq
mbnational <- get_variable(pseq, "nationality" )

# this collapses the two varialbes into a new variable AB in this case. The command for multiple changes is fct_collapse(x, AB = c("A","B"), DE = c("D","E"))
mndmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

pdmbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

admbnational <- fct_collapse(mbnational, LOW = c("Scandinavia","EasternEurope") , MEDIUM = c("SouthEurope","CentralEurope", "UKIE"), HIGH = c("US") )

# reorder
mndmbnational <- factor(mndmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
pdmbnational <- factor(pdmbnational, levels = (c("LOW", "MEDIUM", "HIGH")))
admbnational <- factor(admbnational, levels = (c("LOW", "MEDIUM", "HIGH")))

# creates a new variable in the phyloseq called mnd
sample_data(pseq)$mnd = mndmbnational
sample_data(pseq)$pd = pdmbnational
sample_data(pseq)$ad = admbnational

# checks that it has worked.
#get_variable(pseq, "mnd")

# Clostridum grouping
```{r Clostridium grouping}

clostridium <- subset_taxa(pseq, Phylum =="Clostridium cluster I" | Phylum =="Clostridium cluster XI"| Phylum =="Clostridium cluster XVI"  | Phylum =="Clostridium cluster III"  | Phylum =="Clostridium cluster XIII" | Phylum =="Clostridium cluster XVII"| Phylum =="Clostridium cluster IV" | Phylum =="Clostridium cluster XVIa" | Phylum =="Clostridium cluster XVIII" | Phylum =="Clostridium cluster IX" | Phylum =="Clostridium cluster XV" | Phylum =="Uncultured Clostridiales")

pseq.new <- subset_taxa(pseq, Phylum =="Actinobacteria" | Phylum =="Bacilli"| Phylum =="Proteobacteria"  | Phylum =="Verruccomicrobia"  | Phylum =="Bacteroidetes" | Phylum =="Asteroleplasma"| Phylum =="Spirochaetes" | Phylum =="Fusobacteria" | Phylum =="Cyanobacteria" | Phylum =="Uncultured Mollucutes")

## pseq summary
```{r pseq}
pseq
```

```{r List of genera in phylum}
# List of Genera in the Bacteroideted Phylum
taxa <- map_levels(NULL, "Phylum", "Genus", pseq)$Bacteroidetes
print(taxa)
```

```{r MND table summary}
table(sample_data(pseq)$nationality)
table(sample_data(pseq)$mnd)
table(sample_data(pseq)$pd)
table(sample_data(pseq)$ad)
```

## Summary REMOVE?
```{r summary}

# Summarise contents of phyloseq
summarize_phyloseq(pseq)
# Sparsity is how populated is the data with zeros
kable(head(tax_table(pseq)))
tax_table(pseq)

# Check the data
print(pseq)

# Display variables from the mapping file
sample_variables(pseq)

# Total number of taxa in the entire data
ntaxa(pseq)

# Taxonomic ranks
rank_names(pseq)

## Check OTUs/Phylums
Prune(remove) taxa because we want to remove OTUs/taxa that aren't in any of the samples
Don't need to prune this taxa because OTUs are found in all samples.
Check this everytime you use the subset_samples function because of the pseq object being filtered. 
```{r check OTUs}
# Check if any OTUs are not present in any samples
any(taxa_sums(pseq) == 0)
# FALSE, therefore, OTUs are found in all samples and we don't need to prune the taxa. If the answer is TRUE, some OTUs are not present in any samples. This is usually the case when data is subset to remove some samples. OTUs unique to those sample are not removed along with the samples. Therefore, it is important to check this everytime the phyloseq object is filtered for samples using `subset_samples` function.

head(meta(pseq))

## ALPHA DIVERSITY
Diversity: Shannon, Simpson
Richness: Chao, ACE

## Prepare Data for Alpha Diversity Visualisation
```{r prepare data for alpha diversity visualisation}

# Prepare data for vizualisation
# Alpha diversity for the entire data set
png("alpha.png")
tab <- diversities(pseq, index = "all")
stargazer(tab, type = "text",summary = NULL)
kable(head(tab))
plot(tab)
dev.off()

# Get metadata
pseq.meta <- meta(pseq)
kable(head(pseq.meta))
stargazer(pseq.meta, type = "text",summary = NULL)

# Add the diversity table to metadata
pseq.meta$Shannon <- tab$shannon 
pseq.meta$InverseSimpson <- tab$inverse_simpson
pseq.meta$fisher <- tab$fisher
pseq.meta$chao1 <- tab$chao1

## Nationality Violin Plot
```{r nationality violin plot}

# Alpha diversity is the species richness of a place
# High alpha diversity = high no. of species

pseq.meta$nationality <- factor(pseq.meta$nationality, levels = c("Scandinavia","EasternEurope","CentralEurope", "SouthEurope", "UKIE", "US"))

# Violin plot
# the number in the palettes must equal the number of groups in nationlaity.pairs
p1 <- ggviolin(pseq.meta, x = "nationality", y = "Shannon",
      add = "boxplot", fill = "nationality", palette = "paired") +   theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
  print(p1)
ggsave("nation_violin_shannon.png")

p2 <- ggviolin(pseq.meta, x = "nationality", y = "InverseSimpson",
 add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("")
print(p2)
ggsave("nation_violin_simpson.png")

p3 <- ggviolin(pseq.meta, x = "nationality", y = "fisher",
 add = "boxplot", fill = "nationality", palette = "paired") + theme(axis.text.x = element_text(angle = 90)) + theme(legend.position="none") + xlab("") + ylab("Fisher")
print(p3)
ggsave("nation_violin_fisher.png")

plot_grid(p1, p2, p3, align = "h" , ncol = 2, nrow = 2)
ggsave("nation_violin.png")

```{r}

p <- plot_richness(pseq, x="nationality", measures=c("Chao1", "Shannon", "Simpson")) 
p + geom_boxplot()
ggsave("alpha_nationality.png")

## MND Violin Plot
```{r mnd violin plot}

# Add the diversity table to metadata
pseq.meta$Shannon <- tab$shannon
pseq.meta$InverseSimpson <- tab$inverse_simpson

# Compare differences in Shannon index between mnd group of the study subjects
pseq.meta$mnd <- factor(pseq.meta$mnd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

pseq.meta$pd <- factor(pseq.meta$pd, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

pseq.meta$ad <- factor(pseq.meta$ad, levels = c("LOW","MEDIUM","HIGH"),  exclude = "NA")

# Create a list of pairwise comaprisons
mnd <- levels(pseq.meta$mnd) # get the variables
pd <- levels(pseq.meta$pd) # get the variables
ad <- levels(pseq.meta$ad) # get the variables

# Make a pairwise list that we want to compare.
mnd.pairs <- combn(seq_along(mnd), 2, simplify = FALSE, FUN = function(i)mnd[i])
mnd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

pd.pairs <- combn(seq_along(pd), 2, simplify = FALSE, FUN = function(i)pd[i])
pd.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

ad.pairs <- combn(seq_along(ad), 2, simplify = FALSE, FUN = function(i)ad[i])
ad.pairs <- list( c("LOW","MEDIUM"), c("HIGH","LOW"), c("MEDIUM","HIGH"))

# Shannon
# the number in the palettes must equal the number of groups in bmi.pairs
p1 <- ggviolin(pseq.meta, x = "mnd", y = "Shannon", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
p1 <- p1 + stat_compare_means(comparisons = mnd.pairs, paired = FALSE) + xlab("MND Incidence") + theme(legend.position="none")
print(p1)
ggsave("mnd_violin__shannon_stats.png")

# Inverse Simpson
p2 <- ggviolin(pseq.meta, x = "mnd", y = "InverseSimpson", add = "boxplot", fill = "mnd", palette = c("yellow", "orange", "red"))
p2 <- p2 + stat_compare_means(comparisons = mnd.pairs) 
print(p2)
ggsave("mnd_violin_simpson.png")

## BETA DIVERSITY
Beta-diversity: Measures for differences between samples from different groups to identify if there are differences in the overall community composition and structure.
WE want to: quantify community divergence within a given sample set
Divergence of a given sample can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity.

```{r Phylum NMDS}

png("Phylum_NMDS.png")
pseq.ord <- ordinate(pseq, "NMDS", "bray")
p1 = plot_ordination(pseq, pseq.ord, type = "taxa", color = "mnd", title = "Taxa by Phylum")
p1 + facet_wrap(~Phylum, 3)
dev.off()

 The low group tends to have smaller values, indicating that the samples are more similar to the group mean, and is less heterogeneous (has smaller spread / is more homogeneous):

# VISUALISE MICORBIOME VARIATION
```{r convert to compositional data}

# Convert to compositional data
pseq.rel <- microbiome::transform(pseq, "compositional")
otu <- abundances(pseq.rel)
meta <- meta(pseq.rel)

NMDS/Bray
Bray-Curtis dissimilarity does not consider phylogenetic relationships between OTUs. There are several distance methods and a list can be obtained by typying ?distanceMethodList in the console pane.
```{r}

pseq.ord <- ordinate(pseq.rel, "NMDS", "bray")

p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
plot(p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
ggsave("NMDStMND.png")

p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, nrow = 2) + labs(color = "MND Prevalence")
ggsave("NMDS_facetMND.png")

p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
plot(p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
ggsave("NMDStpd.png")

p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, nrow = 2) + labs(color = "PD Prevalence")
ggsave("NMDS_facetpd.png")

p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
plot(p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
ggsave("NMDStad.png")

p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, nrow = 2) + labs(color = "AD Prevalence")
ggsave("NMDS_facetad.png")

MDS ordination on Bray-Curtis distance
Performs principal coordinate analysis (also called principle coordinate decomposition, multidimensional scaling (MDS), or classical scaling) of a distance matrix (Gower 1966), including two correction methods for negative eigenvalues. See pcoa for further details.
```{r}

pseq.ord <- ordinate(pseq.rel, "MDS", "bray")

p1 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "mnd")
plot(p1 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence"))
ggsave("MDS.png")

p1 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~mnd, 3) + labs(color = "MND Prevalence")
ggsave("NMDS_facet.png")

p2 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "pd")
plot(p2 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence"))
ggsave("MDSpd.png")

p2 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~pd, 3) + labs(color = "PD Prevalence")
ggsave("NMDS_facetPD.png")

p3 = plot_ordination(pseq.rel, pseq.ord, type = "sample", color = "ad")
plot(p3 + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence"))
ggsave("MDSad.png")

p3 + scale_colour_manual(values = c("yellow", "orange", "red", "grey")) + facet_wrap(~ad, 3) + labs(color = "AD Prevalence")
ggsave("NMDS_facetAD.png")

PCoA for compositional data with Bray-Curtis distances
```{r}

p <- plot_landscape(pseq.rel, method = "NMDS", distance = "bray", colour = "mnd", size = 3) 
p2 <- p + scale_color_brewer(palette = "Dark2")+ scale_fill_gradient(low = "#e0ecf4", high = "#6e016b")

p1 <- plot_landscape(pseq.rel, method = "MDS", distance = "bray", col = "mnd", size = 3) 
print(p1)

```{r PCoA Bray}

# Pick core taxa
pseq.core <- core(pseq.rel, detection = 5/100, prevalence = 5/100)

# PCoA for compositional data with Bray-Curtis distances
png("PCoA_bray.png")
p <- plot_landscape(microbiome::transform(pseq.core, "compositional"),
                      method = "PCoA", distance = "bray") +
       labs(title = paste("PCoA / Compositional / Bray-Curtis"))
print(p)
dev.off()

t-SNE is a popular new ordination technique
```{r}

tsne_res <- tsne_phyloseq(pseq, distance='bray', perplexity = 5, verbose=1, rng_seed = 3901)
#MND
tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
  color = 'mnd', title='t-SNE (Bray)')
tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "MND Prevalence")
ggsave("mnd_t_sne.png")
# PD
tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
  color = 'pd', title='t-SNE (Bray)')
tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "PD Prevalence")
ggsave("pd_t_sne.png")

# AD
tsne_p <- plot_tsne_phyloseq(pseq, tsne_res,
  color = 'ad', title='t-SNE (Bray)')
tsne_p + scale_colour_manual(values = c( "yellow", "orange", "red", "grey")) + labs(color = "AD Prevalence")
ggsave("ad_t_sne.png")

Divergence of a given sample set can be quantified as the average dissimilarity of each sample from the group mean; the dissimilarity can be quantified by beta diversity, for instance. This was applied in group-level comparisons for instance in Salonen et al. ISME J 2014 (they focused on homogeneity using inverse correlation, whereas here we focus on divergence using correlation but the measure is essentially the same).
```{r Calculate group divergences}

# Use these to compare microbiota divergence within each group

#MND
mnd.p.low <- subset_samples(pseq, mnd == "LOW")
mnd.p.low.d <- divergence(mnd.p.low)

mnd.p.med <- subset_samples(pseq, mnd == "MEDIUM")
p.med.d <- divergence(mnd.p.med)

mnd.p.high <- subset_samples(pseq, mnd == "HIGH")
p.high.d <- divergence(mnd.p.high)

mnd.p <- boxplot(list(Low = mnd.p.low.d, Medium = mnd.p.med.d, High = mnd.p.high.d))
plot(mnd.p)

# PD

pd.p.low <- subset_samples(pseq, pd == "LOW")
pd.p.low.d <- divergence(pd.p.low)

pd.p.med <- subset_samples(pseq, pd == "MEDIUM")
p.med.d <- divergence(pd.p.med)

pd.p.high <- subset_samples(pseq, pd == "HIGH")
p.high.d <- divergence(pd.p.high)

pd.p <- boxplot(list(Low = pd.p.low.d, Medium = pd.p.med.d, High = pd.p.high.d))
plot(pd.p)

# AD

ad.p.low <- subset_samples(pseq, ad == "LOW")
ad.p.low.d <- divergence(ad.p.low)

ad.p.med <- subset_samples(pseq, ad == "MEDIUM")
p.med.d <- divergence(ad.p.med)

ad.p.high <- subset_samples(pseq, ad == "HIGH")
p.high.d <- divergence(ad.p.high)

ad.p <- boxplot(list(Low = ad.p.low.d, Medium = ad.p.med.d, High = ad.p.high.d))
plot(ad.p)

## PERMANOVA significance test for group-level differences
Evaluate whether the MND grouping has a significant effect on overall gut microbiota composition

```{r Redefine data}

# Convert to compositional data
# MND
mnd.pseqLvH <- merge_phyloseq(mnd.p.low,mnd.p.high)
mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)

mnd.pseq.rel <- microbiome::transform(mnd.pseqLvH, "compositional")
mnd.otu <- abundances(mnd.pseq.rel)
mnd.meta <- meta(mnd.pseq.rel)

# PD

pd.pseqLvH <- merge_phyloseq(pd.p.low,pd.p.high)
pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)

pd.pseq.rel <- microbiome::transform(pd.pseqLvH, "compositional")
pd.otu <- abundances(pd.pseq.rel)
pd.meta <- meta(pd.pseq.rel)

# AD

ad.pseqLvH <- merge_phyloseq(ad.p.low,ad.p.high)
ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)

ad.pseq.rel <- microbiome::transform(ad.pseqLvH, "compositional")
ad.otu <- abundances(ad.pseq.rel)
ad.meta <- meta(ad.pseq.rel)

```{r PERMANOVA}

# samples x species as input
library(vegan)

# MND
mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
               data = meta, permutations=999, method = "bray")
# P-value
print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
summary(mnd.permanova$aov.tab)

# PD
pd.permanova <- adonis(t(pd.otu) ~ pd,
               data = meta, permutations=999, method = "bray")
# P-value
print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
summary(pd.permanova$aov.tab)

# AD
ad.permanova <- adonis(t(ad.otu) ~ ad,
               data = meta, permutations=999, method = "bray")
# P-value
print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
summary(ad.permanova$aov.tab)

Check that variance homogeneity assumptions hold (to ensure the reliability of the results):Check that variance homogeneity assumptions hold (to ensure the reliability of the results):
```{r ANOVA}
#MND
mnd.dist <- vegdist(t(mnd.otu))
mnd.anov <- anova(betadisper(mnd.dist, meta$mnd))
summary(mnd.anov)
tidy(mnd.anov)

#PD
pd.dist <- vegdist(t(pd.otu))
pd.anov <- anova(betadisper(pd.dist, meta$pd))
summary(pd.anov)
tidy(pd.anov)

#AD
ad.dist <- vegdist(t(ad.otu))
ad.anov <- anova(betadisper(ad.dist, meta$ad))
summary(ad.anov)
tidy(ad.anov)

Show coefficients for the top taxa separating the groups
```{r top taxa L vs H}

# MDN
mnd.coef <- coefficients(mnd.permanova)["mnd1",]
mnd.top.coef <- coef[rev(order(abs(mnd.coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(mnd.top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus") 
ggsave("mndLvH.png")

# PD
pd.coef <- coefficients(pd.permanova)["pd1",]
pd.top.coef <- coef[rev(order(abs(pd.coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(pd.top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus") 
ggsave("pdLvH.png")

# PD
ad.coef <- coefficients(ad.permanova)["ad1",]
ad.top.coef <- coef[rev(order(abs(ad.coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(ad.top.coef), col = rainbow(20), horiz = T, las = 1, main = "Top Genus") 
ggsave("adLvH.png")

```{r L vs M}

#MND
mnd.pseqLvM <- merge_phyloseq(mnd.p.low,mnd.p.med)
mnd.pseq.rel <- microbiome::transform(mnd.pseqLvM, "compositional")
mnd.otu <- abundances(mnd.pseq.rel)
mnd.meta <- meta(mnd.pseq.rel)

#PD
pd.pseqLvM <- merge_phyloseq(pd.p.low,pd.p.med)
pd.pseq.rel <- microbiome::transform(pd.pseqLvM, "compositional")
pd.otu <- abundances(pd.pseq.rel)
pd.meta <- meta(pd.pseq.rel)

#AD
ad.pseqLvM <- merge_phyloseq(ad.p.low,ad.p.med)
ad.pseq.rel <- microbiome::transform(ad.pseqLvM, "compositional")
ad.otu <- abundances(ad.pseq.rel)
ad.meta <- meta(ad.pseq.rel)

# MND
library(vegan)
mnd.permanova <- adonis(t(mnd.otu) ~ mnd,
               data = meta, permutations=999, method = "bray")
# P-value
print(as.data.frame(mnd.permanova$aov.tab)["mnd", "Pr(>F)"])
mnd.dist <- vegdist(t(mnd.otu))
mnd.anov <- anova(betadisper(mnd.dist, mnd.meta$mnd))
summary(mnd.anov)
tidy(mnd.anov)
mnd.coef <- coefficients(mnd.permanova)["mnd1",]
mnd.top.coef <- coef[rev(order(abs(mnd.coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(mnd.top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
ggsave("mndLvM.png", width = 20, height = 20, units = "cm")

# PD
library(vegan)
pd.permanova <- adonis(t(pd.otu) ~ pd,
               data = meta, permutations=999, method = "bray")
# P-value
print(as.data.frame(pd.permanova$aov.tab)["pd", "Pr(>F)"])
pd.dist <- vegdist(t(pd.otu))
pd.anov <- anova(betadisper(pd.dist, pd.meta$pd))
summary(pd.anov)
tidy(pd.anov)
pd.coef <- coefficients(pd.permanova)["pd1",]
pd.top.coef <- coef[rev(order(abs(pd.coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(pd.top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
ggsave("pdLvM.png", width = 20, height = 20, units = "cm")

# AD
library(vegan)
ad.permanova <- adonis(t(ad.otu) ~ ad,
               data = meta, permutations=999, method = "bray")
# P-value
print(as.data.frame(ad.permanova$aov.tab)["ad", "Pr(>F)"])
ad.dist <- vegdist(t(ad.otu))
ad.anov <- anova(betadisper(ad.dist, ad.meta$ad))
summary(ad.anov)
tidy(ad.anov)
ad.coef <- coefficients(ad.permanova)["ad1",]
ad.top.coef <- coef[rev(order(abs(ad.coef)))[1:20]]
par(mar = c(3, 14, 2, 1))
barplot(sort(ad.top.coef), col = rainbow(20), horiz = T, las = 1, font.axis = 3, main = "Top Genus")
ggsave("adLvM.png", width = 20, height = 20, units = "cm")

## MICROBIOME COMPOSITION

```{r Phylum #}

# Show available ranks in the dataset
rank_names(pseq)

# Create table, number of features for each phyla
table(tax_table(pseq)[, "Phylum"], exclude = NULL)

```
This shows a few phyla for which only one feature was observed. Those may be worth filtering.

```{r Total Phylum abundnace bar}

png("Phylum_abundance.png")
p<- plot_bar(pseq,x="Phylum",fill = "Phylum") +geom_bar(stat="identity")

phylumGlommed = tax_glom(pseq, "Phylum")
plot_bar(phylumGlommed, x="Phylum", fill ="Phylum") +geom_bar(stat="identity")

pd <- p$data %>% 
    as_tibble %>%
    mutate(Genus = as.character(Genus))
genus_abun <- pd %>%
    group_by(Genus) %>%
    summarize(Abundance = sum(Abundance)) %>%
    arrange(Abundance)
genus_levels <- genus_abun$Genus
pd0 <- pd %>%
    mutate(Genus = factor(Genus, genus_levels))
ggplot(pd0, aes(x = Sample, y = Abundance, fill = Genus)) + 
    geom_bar(stat = "identity")
dev.off()

```{r Phylum prevalence table}

prev.otu <- plot_taxa_prevalence(pseq, "Phylum")
prev.otu + theme(legend.position="none")
ggsave("phylum_prevalence.png")

tab <- plot_taxa_prevalence(pseq, "Phylum", detection = 5)
tab + theme(legend.position="none")
ggsave("Phylum_prevalence_detection5.png")

theme(axis.text.x = element_text(angle = 90)
+ guide_italics
  ggtitle("Relative abundance")  + 
```{r Relative abundance}

pseq.ph <- aggregate_taxa(pseq.new, "Phylum")

pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
  theme(legend.position = "bottom") + theme_bw() + 
  theme(axis.text.x=element_blank()) + 
  theme(legend.title = element_text(size=18)) + scale_fill_brewer(palette = "Paired")
print(plot.comp.rel)
ggsave("phylum_relative_abundance11.png")

plot.comp.rel + theme(legend.position="none")
ggsave("phylum_relative_abundance12.png")

guide_italics
ggtitle("Relative abundance") +  
```{r}

clos.ph <- aggregate_taxa(clostridium, "Phylum")

clos.ph.rel <- microbiome::transform(clos.ph, "compositional")

plot.comp.rel <- plot_composition(clos.ph.rel, x.label = "sample") + 
  theme(legend.position = "bottom", ) + theme_bw() + 
  theme(axis.text.x=element_blank()) + 
  theme(legend.title = element_text(size=18))
plot.comp.rel + scale_fill_brewer(palette = "Paired")

plot.comp.rel <- plot_composition(pseq.ph.rel, x.label = "sample") +
 theme(legend.position = "bottom") + theme_bw() +
 theme(axis.text.x=element_blank()) +
 theme(legend.title = element_text(size=18),legend.text = element_text( face="italic") ) + scale_fill_brewer(palette = "Paired") + guides(fill=guide_legend("my awesome title"))

ggsave("phylum_relative_abundance_clos.png")

plot.comp.rel + scale_fill_brewer(palette = "Paired") + theme(legend.position="none")
ggsave("phylum_relative_abundance_clos2.png")

```{r Phylum Bar Plot}

levels(pseq.meta$mnd)

pseq.ph <- aggregate_taxa(pseq, "Phylum")
pseq.ph.rel <- microbiome::transform(pseq.ph, "compositional")
#MND
plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "mnd", x.label = "MND Incidence", plot.type = "barplot")
plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("mndphylum_abundance_bar.png")
#PD
plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "pd", x.label = "PD Incidence", plot.type = "barplot")
plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("pdphylum_abundance_bar.png")
#AD
plot.comp.rel <- plot_composition(pseq.ph.rel, average_by = "ad", x.label = "AD Incidence", plot.type = "barplot")
plot.comp.rel + scale_fill_viridis(discrete = TRUE, option = "plasma") 
ggsave("adphylum_abundance_bar.png")
```

```{r pseq.new bar plot}

p.new <- aggregate_taxa(pseq.new, "Phylum")

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/globalmicrobiome')

@@

setwd("~/globalmicrobiome")
mnd.incidence <- read_csv("mnd.csv",col_names = TRUE)
mnd.incidence <- as.data.frame(mnd.incidence)

mb.gather <- function (x) {
    mb.group <- data.frame("Location" = c("Central Europe","Eastern Europe","Scandinavia", "Southern Europe","UKIE","USA"), "Mean" = c(1:6)) 
  c.europe <- filter(x,Location=="Belgium"|Location=="Denmark"|Location=="Germany"|Location=="The Netherlands")
    e.europe <- filter(mnd.incidence,Location=="Poland")
    scan <- filter(mnd.incidence,Location=="Finland"|Location=="Norway"|Location=="Sweden")
    s.europe <- filter(mnd.incidence,Location=="France"|Location=="Italy"|Location=="Serbia"|Location=="Spain")
    ukie <- filter(mnd.incidence,Location=="United Kingdom"|Location=="Ireland")
    usa <- filter(mnd.incidence,Location=="United States")
    mb.group[1,2] <- mean(c.europe$Value)
    mb.group[2,2] <- mean(e.europe$Value)
    mb.group[3,2] <- mean(scan$Value)
    mb.group[4,2] <- mean(s.europe$Value)
    mb.group[5,2] <- mean(ukie$Value)
    mb.group[6,2] <- mean(usa$Value)
    return(mb.group)
    }

temp <- mb.gather(mnd.incidence)

@@
